Reproducible Research: Peer Assessment 1
========================================

## Loading and preprocessing the data

Make sure packages dplyr and ggplot2 are installed and the data file activity.csv is in the working directory.

Load the packages:
```{r,message=FALSE}
library(dplyr)
library(ggplot2)
```

Load the data:
```{r}
data <- read.csv("activity.csv", sep = ",", header = TRUE)
```

Filter out missing data:
```{r}
data_nona <- filter(data, !is.na(steps))
```


## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day:
```{r}
data_nona_tot <- select(data_nona, date, steps) %>%
    group_by(date) %>% summarise_each(funs(sum))
```

Make a histogram of the total number of steps taken each day:
```{r}
ggplot(data_nona_tot, aes(x = steps)) +
     geom_histogram(binwidth = 1000, color = "black", fill = "white") +
     xlab("Steps per day") +
     ylab("Days")
```

Calculate the median total number of steps taken each day:
```{r}
median(data_nona_tot$steps)
```

Calculate the mean total number of steps taken each day:
```{r}
round(mean(data_nona_tot$steps), digits = 0)
```

## What is the average daily activity pattern?
Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):
```{r}
data_avg <- select(data_nona, interval, steps) %>% 
    group_by(interval) %>% summarise_each(funs(mean)) %>%
    mutate(steps = round(steps, digits = 0))
ggplot(data_avg, aes(interval, steps)) +
    geom_line() +
    xlab("Interval") +
    ylab("Number of steps")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, message=FALSE}
interval_maxsteps <- filter(data_avg, steps == max(data_avg$steps))
interval_maxsteps
```

The answer is `r interval_maxsteps[1]`.

## Imputing missing values

Collect missing data:
```{r}
data_na <- filter(data, is.na(steps))
```

There are `r nrow(data_na)` rows with missing values. 

Let's use the mean for the 5-minute interval to fill in all of the missing values in the dataset.

```{r}
data_impute <- merge(select(data_na, date, interval), data_avg, 
    by.x = "interval", by.y = "interval")
```

Create a new dataset that is equal to the original dataset but with the missing data filled in:
```{r}
data_new <- rbind(data_nona, data_impute) %>% arrange(date, interval)
```

Make a histogram of the total number of steps taken each day:
```{r}
data_new_tot <- select(data_new, date, steps) %>% 
    group_by(date) %>% summarise_each(funs(sum))
ggplot(data_new_tot, aes(x = steps)) +
    geom_histogram(binwidth = 1000, color = "black", fill = "white") +
    xlab("Steps per day") +
    ylab("Days")
```

Calculate the median total number of steps taken each day:
```{r}
median(data_new_tot$steps)
```

Calculate the mean total number of steps taken each day:
```{r}
round(mean(data_new_tot$steps), digits = 0)
```

In comparison to the dataset before imputing, the medians differ and the means are equal. The median is slightly lower after applying the imputing strategy.

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
data_wkd <- mutate(data_new, day_type = as.factor(ifelse(weekdays(as.Date(date)) %in% 
    c("Saturday", "Sunday"), "weekend", "weekday")))
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
data_wkd_avg <- select(data_wkd, day_type, interval, steps) %>% 
    group_by(day_type, interval) %>% summarise_each(funs(mean)) %>%
    mutate(steps = round(steps, digits = 0))
ggplot(data_wkd_avg, aes(interval, steps)) +
    geom_line() +
    facet_wrap(~day_type, ncol = 1) + 
    xlab("Interval") +
    ylab("Number of steps")
```

From the plot, we can see more steps in weekdays between an interval around 500 and an interval around 955. On the other hand, weekends see more activities after an interval around 955 and before an interval around 2130.